---
title: "Hypothesis_testing1"
author: "Diana Nduku"
date: "2023-10-08"
output: html_document
---



```{r}
## importing the dataset
data= read.csv("C:/Users/USER/OneDrive/Data Science books/datasets/titanic dataset/tested.csv")

head(data)

```

# DATA CLEANING

```{r}

## dropping the passenger ID, cabin, ticket and names column since they are not useful
df=data[,-c(1,4,9,11)]
head(df)
```



```{r}
##checking for duplicates
duplicates=(data[duplicated(data)])
sum_duplicates=sum(duplicates)
sum_duplicates

#no duplicate values

##checking for missing values
missing_values <- colSums(is.na(df))
missing_values

## missing values are only in age and fare category

## replace missing fare value with mean fare price
mean_fare = mean(df$Fare, na.rm = TRUE)
df$Fare[is.na(df$Fare)] = mean_fare

## replace age with the mean age for each class

head(df)

# Calculate the mean age for each Pclass using tapply()
mean_age_by_pclass =tapply(df$Age, df$Pclass, FUN = mean, na.rm = TRUE)

## replacing the missing values
df$Age = ifelse(is.na(df$Age), mean_age_by_pclass[as.character(df$Pclass)], df$Age)

## recheckingg missing values
missing_values = colSums(is.na(df))
missing_values

## no missing values
```

# DATA EXPLORATION

```{r}
#Descriptive Statistics
summary(df)

#correlation matrix for the numeric variables
#selecting numeric variables
numeric_variables=sapply(df, is.numeric)
numeric_data=df[, numeric_variables]
correlation_matrix=cor(numeric_data)
correlation_matrix
```

#### There is a moderately strong relationship between class and age given class' age mean was used to fill the missing NA values for age
#### There is also a moderately strong negative relationship between passenger class and Fare prices.The higher the price, the lower the class number(1st class being the lowest and 3rd the highest)





```{r}
#variable frequencies

sex_frequencies=table(df$Sex)
sex_frequencies
barplot(sex_frequencies)
```

#### There were more male passengers 



```{r}
class_frequencies=table(df$Pclass)
class_frequencies
barplot(class_frequencies)
```

#### Most passengers were in 3rd class and the least in 2nd class

```{r}
survived_frequencies=table(df$Survived)
survived_frequencies
barplot(survived_frequencies)
```

##### 0= did not survive, 1= survived. Majority of the passengers did not survive.


```{r}
embarked_frequencies=table(df$Embarked)
embarked_frequencies
barplot(embarked_frequencies)
```

#### Majority of passengers embarked at station S, followed by C and the least at Q




```{r}
#place age into categories and fing the frequencies
age_categories <- cut(df$Age, 
                      breaks = c(0, 18, 30, 50, max(df$Age, na.rm = TRUE)),
                      labels = c("0-18", "19-30", "31-50", "51+"),
                      include.lowest = TRUE,
                      right = FALSE)
age_frequencies=table(age_categories)
age_frequencies
barplot(age_frequencies)
```

#### Majority of the passengers were aged between 19-30. 0-18 had the least number of passengers
```{r}
##place fare into categories and find frequencies
fare_categories <- cut(df$Fare,
                       breaks = c(0, 10, 30, 50, 100, max(df$Fare, na.rm = TRUE)),
                       labels = c("0-10", "11-30", "31-50", "51-100", "101+"),
                       include.lowest = TRUE,
                       right = FALSE)

fare_frequencies=table(fare_categories)
fare_frequencies
barplot(fare_frequencies)

boxplot(df$Fare)
```

##### Majority of the passenger's fare ranged from 0-30.
##### There is one outlier: one passenger paid a significantly higher amount of fare


# HYPOTHESIS TESTING
## HYPOTHESIS 1

### Ho:There is no statistical and significant difference between the ages of survivors vs non-survivors
### H1:The mean age of survivors is less than the mean age for non-survivors(A higher number of young people survived)

```{r}

library(dplyr)
survivors=filter(df, df$Survived==1)
head(survivors)

non_survivors=filter(df,df$Survived==0)
head(non_survivors)

```
```{r}
##performing t-test
t_test_age<- t.test(survivors$Age, non_survivors$Age, 
                    alternative='less',
                    paired=FALSE)
t_test_age

```
### Interpretation:The p-value is 0.541, which is larger than 0.05. Therefore, we fail to reject the null hypothesis
### Conclusion: There is no statistical and significant difference in the ages of survivors vs non survivors.The differences are due to random variation

## HYPOTHESIS 2
### Ho:There is no association between survival rate  and passenger class
### H1: There is a statistical and significant association between survival rate and passenger class

```{r}
#contingency table
contingency_table=table(df$Survived, df$Pclass)
contingency_table

#chi-square test
chisq.test(contingency_table)

```

### Interpretation: Since the p-value  is 0.03, which is below 0.05, we have sufficient evidence to reject the null hypothesis
### Therefore we conclude that there is a significant association between passenger class and survival rate




```{r}
##Determining which passenger class had a higher survival rate
survival_percentages = with(df, prop.table(table(Pclass, Survived), margin = 1) * 100)
survival_percentages

```
### Conclusion: Passengers in 1st class had a higher survival rate. Survival in 2nd and 3rd class was similar.(0= did not survived, 1= survived)

## HYPOTHESIS 3
### Ho: There is no association between sex and survival rate
### H1: There is a statistical and significant association between sex and survival rate


```{r}

# Convert "Sex" to binary (0 for male, 1 for female)
df$Sex <- ifelse(df$Sex == "male", 0, 1)
head(df)

#contingency table
contingency_table_sex=table(df$Sex,df$Survived)
contingency_table_sex

#performing chi-square test
chisq.test(contingency_table_sex)

```

### Interpretation: The p-value is 2.462e-08 which is significantly smaller than 0.05. We therefore reject the null hypothesis

### Therefore, we can conclude that there is a statistical and significant association between survival rate and sex

```{r}
##Determining which sex had a higher survival rate
sex_survival_percentages = with(df, prop.table(table(Sex, Survived)) * 100)
sex_survival_percentages

```

### male =0, female =1
### 63% of males did not survive, while 36% of females did not survive

# CONCLUSIONS

### We conclude that the biggest indicator for survival rate was sex, followed by passenger class.
### Females had a higher chance of survival. 
### Passengers in 1st class had a higher chance of survival.
### A female passenger in first class was more likely to survive
### A male passenger in 2nd or 3rd class was least likely to survive
### Age did not have an influence on survival rate
